---
title: "Figure 5: Subclone statistics"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, cache=FALSE)

library(dplyr)
library(ade4)
library(sf)
library(taRifx )
library(Rtsne)
library(ggplot2)
library(adephylo)
library(beeswarm)
library(phangorn)
source('funx.R')


set.seed(6)
```



```{r prep, echo=FALSE}
thr=.05 # set to 0 if no calling is required


tumor.summary<-data.frame(
  name=c("C","D","E","F","H","J", "K", "R", "M", "T", "U", "W"),
  reroot=c("A9_4","A4_20", "none", "none", "A11_15", "none", "A21_26", "A7_7", "A3_24", "A7_3", "A5_19", "A6_25"))
```


## Figure 5a: subclone size distribution

```{r clone size v2, echo=FALSE}

w.out<-data.frame(tumor="A",  n=1)

for(nam in tumor.summary$name){

# Process the data to extract the number of subclones
XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
namz<-XoutY[[1]]

t<-table(namz$subclone)
w<-data.frame(tumor=rep(nam, length(t)),  n=as.vector(t))

w.out<-rbind(w.out, w)
}

w.out<-w.out[-1,]



ggplot(w.out, aes(x = n)) +  geom_histogram(color="black", fill="red", binwidth = 1, alpha=0.4)+xlab("Spots per subclone (n)") +scale_x_continuous(breaks=c(1, seq(5,30,5)))+ ylab("Subclones (n)") + theme_classic()+
      theme(axis.text.x=element_text(size=rel(1.8)), 
        axis.text.y=element_text(size=rel(1.8)),
        axis.title.x=element_text(size=rel(2)),
        axis.title.y=element_text(size=rel(2)))


ggsave( file="Fig_5a.pdf", width=5, height=4)


# stats
#median(w.out$n)

```


## Figure 5b: Number of subclones per slide

```{r subclone v1, echo=FALSE}

#########################################
## Subclone distributions for each tumor
#########################################

w.out<-data.frame(tumor="A", slide="A1", n=0)

for(nam in tumor.summary$name){

# Pre-processing I
Xout<-tumor_process(nam)
datG<-Xout[[1]]


###################################
# Pre-processing II
XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
namz<-XoutY[[1]]


##################################
ind<-match(namz$original,datG$spot)
namz$phenotype<-datG$phenotype[ind]
namz$sample<-datG$sample[ind]
namz$slide<-datG$slide[ind]
namz$subclone<-as.factor(as.character(namz$subclone))

if(nam %in% c("C", "J", "H")){namz<-namz[-which(grepl("met", namz$phenotype)), ]}
namz$slide<-droplevels(namz$slide)
w<-data.frame(tumor=nam, slide=levels(namz$slide), n=0)

for(sl in levels(namz$slide)){
  
  ind<-which(namz$slide==sl)
  w$n[which(w$slide==sl)]<-length(unique(namz$subclone[ind]))
  
  
}


w.out<-rbind(w.out, w)
}

w.out<-w.out[-1,]

ggplot(w.out, aes(x = n)) +  geom_histogram(color="black", fill="red", binwidth = 1, alpha=.4)+xlab("Subclones per slide (n)") +scale_x_continuous(breaks=seq(1,10,1))+ ylab("Slides (n)") + theme_classic()+
    theme(axis.text.x=element_text(size=rel(2)), 
        axis.text.y=element_text(size=rel(2)),
        axis.title.x=element_text(size=rel(2)),
        axis.title.y=element_text(size=rel(2)))

ggsave( file="Fig_5b.pdf", width=5, height=4)


# summary stats

#median(w.out$n)
```


## Figure 5c: Number of subclones per tumor

```{r subclone v2, echo=FALSE}

######################
## Subclone distributions for each tumor
######################

clone.n<-data.frame(tumor=tumor.summary$name, n=0)

for(nam in tumor.summary$name){

# Process the data to extrac the number of subclones
XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
namz<-XoutY[[1]]

clone.n$n[clone.n$tumor==nam]<-max(namz$subclone)

}

ggplot(clone.n, aes(x = n)) +  geom_histogram(color="black", fill="red",alpha=0.4, binwidth = 1)+xlab("Subclones per tumor (n)") +scale_x_continuous(breaks=seq(1,10,1))+ ylab("Tumors (n)") + theme_classic()+
  theme(axis.text.x=element_text(size=rel(2)), 
        axis.text.y=element_text(size=rel(2)),
        axis.title.x=element_text(size=rel(2)),
        axis.title.y=element_text(size=rel(2)))


ggsave( file="Fig_5c.pdf", width=5, height=4)

# summary stats

#median(clone.n$n)
        
```


## Figure 5d: Phenotype distribution for invasive carcinomas (Version 1)


```{r phenotype v1`, echo=FALSE}

# Each tumor separately (removing the adenoma K)
w.out<-data.frame(tumor=1, pheno="Mixed", freq=1, n=1)

for(nam in tumor.summary$name[-7]){

###################################
# Pre-processing I
Xout<-tumor_process(nam)
datG<-Xout[[1]]


###################################
# Pre-processing II
XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
namz<-XoutY[[1]]


##################################
ind<-match(namz$original,datG$spot)
namz$phenotype<-datG$phenotype[ind]
namz$sample<-datG$sample[ind]
namz$slide<-datG$slide[ind]

namz$subclone<-as.factor(as.character(namz$subclone))
###################################

n=length(unique(namz$subclone))
v<-data.frame(tumor=nam, clone=levels(namz$subclone), pheno=0)

for(clo in 1:n){
  
  c<-levels(namz$subclone)[clo]
  
  ind<-which(namz$subclone==c)
  
  if(any(grepl("inv", namz$phenotype[ind]) | grepl("met", namz$phenotype[ind])) & any(grepl("sup", namz$phenotype[ind]))){
    
    v$pheno[clo]<-"Mixed"
    
  } else if(all(grepl("inv", namz$phenotype[ind])|grepl("met", namz$phenotype[ind]))){
    
    v$pheno[clo]<-"Invasive/metastatic"
  } else if(all(namz$phenotype[ind]=="super")){
    
    v$pheno[clo]<-"Superficial"
  }
    
}

w<-data.frame(tumor=nam, pheno=names(table(v$pheno)), freq=as.vector(table(v$pheno)/sum(table(v$pheno))), n=as.vector(table(v$pheno)))
w.out<-rbind(w.out,w)

}

w.out<-w.out[-1,]

## All tumors combined

a<-c(sum(w.out$n[w.out$pheno=="Invasive/metastatic"]),
     sum(w.out$n[w.out$pheno=="Superficial"]),
     sum(w.out$n[w.out$pheno=="Mixed"]))

w<-data.frame(tumor="ALL", pheno=c("Invasive/metastatic","Superficial", "Mixed"), freq=a/sum(a), n=a)


w.out<-rbind(w.out,w)


ggplot(w.out, aes(x = tumor, y = freq, fill = pheno)) +  geom_bar(stat = "identity") +ylab("Frequency")+xlab("Carcinoma")+labs(fill = "Subclone phenotype")+scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
      theme(axis.text.x=element_text(size=rel(2.5)), 
        axis.text.y=element_text(size=rel(1.5)),
        axis.title.x=element_text(size=rel(2)),
        legend.text = element_text(size=rel(1.5)),
        legend.title = element_text(size=rel(1.5)),
        axis.title.y=element_text(size=rel(2)))

ggsave( file="Fig_5d.pdf", width=14, height=8)


```



## Figure 5f: Subclone size distribution by early and late

Next, we look at the subclone size distribution by phenotype

```{r clonesize_v3, echo=FALSE}

w.out<-data.frame(tumor="A", subclone="b", timing="early", n=1)

for(nam in tumor.summary$name[-7]){

###################################
# Pre-processing I
Xout<-tumor_process(nam)
datG<-Xout[[1]]

###################################
# Pre-processing II
XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
namz<-XoutY[[1]]
treeMP<-XoutY[[2]]

##################################
ind<-match(namz$original,datG$spot)
namz$phenotype<-datG$phenotype[ind]
namz$sample<-datG$sample[ind]
namz$slide<-datG$slide[ind]

namz$subclone<-as.factor(as.character(namz$subclone))

dr<-distRoot(treeMP)
dr.early<-which(dr<=max(dr)/2)
dr.late<-which(dr>max(dr)/2)
dr[dr.early]<-"early"
dr[dr.late]<-"late"

namz$timing<-dr[match(names(dr),namz$original)]

w<-data.frame(tumor=nam, subclone=levels(namz$subclone), timing="early", n=1)
w$timing<-as.character(w$timing)

for(lev in levels(w$subclone)){

  ind<-which(namz$subclone==lev)

  w$n[which(w$subclone==lev)]<-length(ind)

  w$timing[which(w$subclone==lev)]<-namz$timing[ind[1]]

}

w.out<-rbind(w.out, w)
}

w.out<-w.out[-1,]
w.out$timing<-as.factor(w.out$timing)

# ggplot(w.out, aes(x = n, fill=phenotype, color=phenotype)) +  geom_histogram(alpha = 0.2, binwidth = 1, position='identity')+xlab("Spots per subclone (n)") +scale_x_continuous(breaks=c(1, seq(5,30,5)))+ ylab("Subclones (n)")


ggplot(w.out, aes(x = n, fill=timing, color=timing))+geom_density(alpha=0.4, stat="density")+labs(color="", fill="Timing")+ylab("Density")+xlab("Spots per subclone (n)")+theme_classic()+
  theme(axis.text.x=element_text(size=rel(1.8)),
        axis.text.y=element_text(size=rel(1.8)),
        axis.title.x=element_text(size=rel(2)),
        axis.title.y=element_text(size=rel(2)))

# summary statistics

sum(w.out$timing=="early")
sum(w.out$timing=="late")

ind.early<-which(w.out$timing=="early")
ind.late<-which(w.out$timing=="late")

early.vec<-w.out$n[ind.early]
late.vec<-w.out$n[ind.late]

median(early.vec)
median(late.vec)

wilcox.test(early.vec, late.vec, paired = FALSE)

ggsave( file="Fig_5f.pdf", width=6, height=4)

```


## Figure 5g: normalized distance between neighboring clones

```{r}
#####################################
## Read in
#####################################



names.vec<-c("C","D","E","F","H","J", "K", "R", "M", "T", "U", "W")

c<-(c(1,2,6,4,2, 2)-1)/6
d<-(c(2,3,1,2)-1)/3
e<-(c(2,4,5,2,1,5,5,1)-1)/4
f<-(c(4,1, 3, 2, 3, 3, 2, 3)-1)/3
h<-(c(1,2)-1)/1
k<-(c(2,2,4,4,5,2, 6)-1)/5
m<-(c(2, 3, 4,5, 2, 2, 2, 3, 5)-1)/5
r<-(c(2,2,1)-1)/1
t<-(c(2,2, 1,4)-1)/3
u<-(c(2,2,2)-1)/2
w<-(c(2, 2, 4,5, 2, 4, 3, 1,3)-1)/5



vec<-data.frame(val=c(m,d,k,c,h,e,r, f, t, u,w))

ggplot(vec, aes(x = val)) +  geom_histogram(color="black", fill="red", binwidth = .2, alpha=0.4)+xlab("Adjacent subclone distance") +scale_x_continuous(breaks=c(seq(0,1,.2)))+ ylab("Frequency") + theme_classic()+
      theme(axis.text.x=element_text(size=rel(2.5)), 
        axis.text.y=element_text(size=rel(3)),
        axis.title.x=element_text(size=rel(3)),
        axis.title.y=element_text(size=rel(3)))


ks.test(vec+rnorm(dim(vec)[1], mean=0, sd=0.00000001), "punif", alternative = c("two.sided"))

ggsave( file="Fig_5g.pdf", width=8, height=6)


```



## Figure 5h: Gini-Simpson Index among superficial and invasive clones



```{r combo, echo=FALSE, message=FALSE, warning=FALSE}


tumor.summary<-data.frame(
  name=c("C","D","E","F","H","J", "K", "R", "M", "T", "U", "W"),
  reroot=c("A9_4","A4_20", "none", "none", "A11_15", "none", "A21_26", "A7_7", "A3_24", "A7_3", "A5_19", "A6_25"))



thr=0.05

tumor.summary$super<-0
tumor.summary$inv<-0
tumor.summary$span<-0
tumor.summary$simpson_super<-0
tumor.summary$simpson_inv<-0


#####################################
## 
#####################################



for(nam in tumor.summary$name){
  
  
  ###################################
  # Pre-processing I
  Xout<-tumor_process(nam)
  datG<-Xout[[1]]
  
  ###################################
  # Pre-processing II
  XoutY<-MP_tree_function(thr, tumor.summary, nam, 10)
  namz<-XoutY[[1]]
  dat<-XoutY[[4]]
  
  
  ##################################
  ind<-match(namz$original, datG$spot)
  namz$phenotype<-datG$phenotype[ind]
  namz$sample<-datG$sample[ind]
  namz$slide<-datG$slide[ind]
  namz$subclone<-as.factor(as.character(namz$subclone))
  
  ind<-match(rownames(dat), namz$original)
  labels.subclone<-namz$subclone[ind]
  labels.pheno<-namz$phenotype[ind]
  
  c<-which(namz$phenotype=="inv" | namz$phenotype=="super" | namz$phenotype=="invasive")
  namz<-namz[c,]
  
  a<-which(namz$phenotype=="inv" | namz$phenotype=="invasive")
  down<-unique(namz$subclone[a])
  b<-which(namz$phenotype=="super")
  up<-unique(namz$subclone[b])

  j<-which(tumor.summary$name==nam)
  tumor.summary$super[j]<-length(setdiff(up, down))
  tumor.summary$inv[j]<-length(setdiff(down, up))
  tumor.summary$span[j]<-length(intersect(down,up))
  
  ll<-length(namz$subclone[a])
  f<-table(namz$subclone[a])
  tumor.summary$simpson_inv[j]<-1-sum(f*(f-1)/ll/(ll-1))
  
  ll<-length(namz$subclone[b])
  f<-table(namz$subclone[b])
  tumor.summary$simpson_super[j]<-1-sum(f*(f-1)/ll/(ll-1))
  

}


# remove C (only super), K (adenoma), J (only one clone)

tumor.summary<-tumor.summary[-c(1,7),-2]



```


```{r}


S<-data.frame(coeff=c(tumor.summary$simpson_super,tumor.summary$simpson_inv),
              lab=c(rep("Superficial",10), rep("Invasive",10)))

pdf(file = "Fig_5h.pdf", width = 5, height = 8) # defaults to 7 x 7 inches

beeswarm(coeff ~ lab, data = S, 
         col = c("black", "black"), pch = 16, cex=2,spacing=1.5, cex.main=2,  main="Gini-Simpson Index", xlab="", ylab="",cex.axis=2, cex.lab=2,  ylim=c(0,1))
dev.off()


t.test(tumor.summary$simpson_super, tumor.summary$simpson_inv, paired=TRUE, alternative = "two.sided")

```




  

